# Regression and model validation

*Describe the work you have done this week and summarize your learning.*

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.
```{r}
students2014 <- read.csv("C:/Users/Kalka/OneDrive/Desktop/Random/School/Open_Data/IODS-project/learning2014.csv")
## This dataset consists of information from 166 people on 7 different parameters including their gender,age, and certain learning questionnaire parameters. The point of the questionnaire was to see how the students course results correlate with their learning approaches and attitude towards a certain statistics course. Deep stra and surf correlate to study skill dimensions 
students2014
```
```{r}
# graphical overview of the data can be shown by using ggpairs
library(GGally)
library(ggplot2)
# create an plot matrix with ggpairs()
ggpairs(students2014, lower = list(combo = wrap("facethist", bins = 20)))

# create a regression model with multiple explanatory variables
my_model2 <- lm(Points ~ attitude + stra+surf+deep+Age+gender, data = students2014)

# print out a summary of the model
summary(my_model2)
```

- in this overview you can see the different correlations between points that the student got and the answers they gave. From this we can say that a positive attitude towards statistics correlated positively with points gained from the course. The other factors did not correlate with statistical significance towards the points


- Below is a summary of how this data shows the correlation of attitude age and strategical methods towards a course and the results on a linear regression model. The only one that correlates stronly is once again the attitude, However stra and age also seem to have some correlation but not statistically significant.
-
```{r}
library(ggplot2)

# fit a linear model
my_model <- lm(Points ~ attitude+Age+stra, data = students2014)

# print out a summary of the model
summary(my_model)
```
-Here we run the fitting again removing the variables that had no statistical significances shown by the *** marking and only run it by explaining points with attitude towards the course and we also plot the two variables that show correlation
```{r}
qplot(attitude, Points, data = students2014) + geom_smooth(method = "lm")

# fit a linear model
my_model <- lm(Points ~ attitude, data = students2014)

# print out a summary of the model
summary(my_model)
```


- As you see in the summary of a linear model for this data for every scaled point of attitude our expected points increase by 3.52

-- note however that the R squared value is only 0.19 which means that our model only accounts for 19% of the variability in the data and thus might not be an accurate method to model this data --> in other words points are only 19% explained by course attitude

- Lastly we produce some diagnostic plots as in ex2
```{r}
# create a regression model with multiple explanatory variables
my_model2 <- lm(Points ~ attitude, data = students2014)

# draw diagnostic plots using the plot() function. Choose the plots 1, 2 and 5
my_model2
plot(my_model2,which = c(1,2,5),par(mfrow=c(2,2)))
```
```
- Based on the course book fig 3.4a the Residuals against Fitted values plot shows that our fitted model should be of appropriate quality as there is no clear curvature in our values which would indicate that the model would need transformation.
- for the standardized residuals we see some potential outliers on the left side which may be the reason why our R-squared is so low. Perhaps there are people who get low scores despite having a good attitude score which lower the predictability ?
```{r}
date()
```

Here we go again...
