---
title: "Chapter 6"
author: "Kalle Karvonen"
date: "2023-12-11"
output: html_document
---
# Chapter 6: Analysis of longitudinal data

```{r}
# Data used is based on the data of Kimmo Vehkalahti from the following URL
#https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt
#https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt

ratsl <-read.csv("C:/Users/Kalka/OneDrive/Desktop/Random/School/Open_Data/IODS-project/data/ratsl.csv")


```

```{r}
# Loading in the data might lose categorical variables set as factors  so lets put them again

ratsl$ID<- as.factor(ratsl$ID)
ratsl$Group<-as.factor(ratsl$Group)
head(ratsl)
```

- lets follow the course book chapter 8 analysis but use the rat dataset instead of BPRSL
- The rat dataset describes the weight of 16 rats over 9 weeks of observation, The rats were assigned different diets out of 3 options. Weight is the weight of a rat in grams and time is in days
```{r}
# Initialising some packages that might be needed 
library(ggplot2)
library(readr)
library(dplyr)
library(tidyr)
# Draw the plots
ggplot(ratsl)+geom_line(aes(x=Time,y=Weight,group=ID,color=Group))
```
- a slight increase in weight can be seen in the groups, vut initial weight is much lower in group 1. Largest weight gain seems to be in group 2 

- lets do a standardization and see if it changes anything in our graph
```{r}
# based on Ex6 code
ratsl_std <- ratsl %>%
  group_by(Group) %>%
  mutate(stdweight = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()

# Glimpse the data
glimpse(ratsl_std)
```
```{r}
# Plot again with the new data
ggplot(ratsl_std)+geom_line(aes(x=Time,y=stdweight,group=ID,color=Group))+
  labs(y = "Standardized weight")
```
- Now we can see the pattern of weight growth better which was hard to see previously due to high variation of starting weight.

- Lets plot 
```{r}
# Summary data with mean and standard error of ratsl by treatment and week 
# Function from the ex 6
se <- function(x,n) { sd(x)/sqrt(n) }

ratsl_s <- ratsl %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = se(Weight, length(Weight)) ) %>%
  ungroup()
```
```{r}
glimpse(ratsl_s)
```
```{r}
ggplot(ratsl_s, aes(x = Time, y = mean, color = Group)) +
  geom_line() +
  geom_point(aes(shape=Group), size=2) +
  theme_bw() +
  geom_errorbar(aes(ymin = mean-se, ymax = mean+se), width=2) 
```
- Variation in group 1 is smaller than the others especially group 2 has a large variation which might be due to an outlier group.

- lets do a boxplot to see any potential outliers 

```{r}
ratsl_bp <- ratsl %>%
  group_by(Group, ID) %>%
  summarise(mean=mean(Weight) ) %>%
  ungroup()
```
```{r}
library(ggplot2)
library(ggpubr)
box1 <-ggplot(ratsl_bp, aes(x = ID, y = mean,color=Group,fill=Group)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=3) +
  scale_y_continuous(name = "mean(weight")

box2 <-ggplot(ratsl_bp, aes(x = Group, y = mean, color=Group,fill=Group)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=3,fill="black") +
  scale_y_continuous(name = "mean(weight")

ggarrange(box1,box2, common.legend=T)
```
- Based on this one could say that rats 2 12 and 13 are outliers when compared to their groups.

- Filtering the outliers and trying again

```{r}
ratsl_fltr <- ratsl %>%
  filter(ID != 2 & ID != 12 & ID != 13) %>%
  group_by(Group, ID) %>%
  summarise(mean=mean(Weight) ) %>%
  ungroup()
```
```{r}
library(ggplot2)
library(ggpubr)
box1 <-ggplot(ratsl_fltr, aes(x = ID, y = mean,color=Group,fill=Group)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=3) +
  scale_y_continuous(name = "mean(weight")

box2 <-ggplot(ratsl_fltr, aes(x = Group, y = mean, color=Group,fill=Group)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=3,fill="black") +
  scale_y_continuous(name = "mean(weight")

ggarrange(box1,box2, common.legend=T)
```
- Data seems to be now more uniform 
- Lastly lets fit a linear model and analyze it

```{r}
# We need to filter the base data
ratsl_fltr <- ratsl %>%
  filter(ID != 2 & ID != 12 & ID != 13)

ratsl_lm <- lm(data=ratsl_fltr,Weight~Time+Group)

summary(ratsl_lm)
```
```{r}
anova(ratsl_lm)
```
- there seems to be a very low p-value on all groups so the difference is significant
-R squared of 0.994 says that 99.4% of our variation can be attributed to our variables 

### Chapter 6 part 2 (BPRS data analysis)

-The BPRS data consists of brief psychiatric rating scale values for 40 men
 before and after treatment on different time frames on two different treatment groups
 

```{r}
BPRSL <-read.csv("C:/Users/Kalka/OneDrive/Desktop/Random/School/Open_Data/IODS-project/data/BPRSL.csv")
BPRSL<- as.data.frame(BPRSL)
BPRSL$subject <- as.factor(BPRSL$subject)
BPRSL$treatment <- as.factor(BPRSL$treatment)

summary(BPRSL)
str(BPRSL)
```
- Treatment is a variable with either 1 or 2 which shows the treatment group
- subject is an identifier for the individual
- bprs is the brief psychiatric rating scale
- week is the time  of data collection

```{r}
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject, color=treatment)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```
- both groups have a downwards trend although some Ids show variability

Fitting a linear model

```{r}
# basic linear model
BPRS_lm <- lm(bprs ~ week + treatment, data = BPRSL)
summary(BPRS_lm)
```
- Based on this there would seem to be no difference between treatments and both treatments seem to reduce the symtoms(bprs score)
- how about other model
```{r}
library(lme4)
library(Matrix)
```

```{r}
# random intercept model

BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref)
```
```{r}
#random intercept and slope model
BPRS_ref2 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref2)
```
- finally lets fit a random intercept and slope model

```{r}
anova(BPRS_ref, BPRS_lm)
```
- Based on this there seems to be a statistically significant degrease in bprs score for subjects

- lets test the last one

```{r}
anova(BPRS_ref2,BPRS_ref)
```
- Same can be seen in this but with a smaller P-value

```{r}
#Lets plot our model
Fitted <- fitted(BPRS_ref2)
BPRSL$Fitted <- Fitted

ggplot(BPRSL, aes(x = week, y = Fitted, group = interaction(treatment, subject))) +
  geom_line(aes(linetype = treatment, colour=treatment)) +
  scale_x_continuous(name = "weeks")+
  scale_y_continuous(name = "Fitted BPRS") +
  theme(legend.position = "top")  +
  ggtitle("Fitted models for BPRS data")


```
- From this it would seem that both treatments reduce the symptoms and neither seems to be better than the other.

#### last assignment done
